A Three Steps Workflow for Simulating Dispersal Trajectories and Assessing Landscape Connectivity: A Case Study on the Endangered African Wild Dog

### 2022-01-05 Chapter_1/04_Manuscript/Manuscript.tex
Many updates

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{Step by Step: Using Step Selection Functions to Simulate Dispersal and
Assess Landscape Connectivity}

% \title{Step by Step: The Utility of Dispersal Simulations to Assess Landscape
% Connectivity}

@@

% Current Date
\date{\today}

@@

\textsuperscript{2} Botswana Predator Conservation, Private Bag 13, Maun,
Botswana.

@@

\textbf{Running Title:} Release the Dogs! Simulating Wild Dog Dispersal to
Assess Landscape Connectivity

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
%------------------------------------------------------------------------------
% Short Version
%------------------------------------------------------------------------------
\textbf{Short Version }
The ability to disperse is contingent on a sufficient degree of landscape
connectivity, which is why the identification and preservation of movement
corridors that promote connectivity has become a task of extraordinary
importance. Currently, ecologists rely on least-cost analysis or circuit
theory to investigate connectivity, albeit both methods make several
assumptions that are hardly met in reality. To address these issues,
simulations from individual-based movement models have been proposed, yet a
unified framework to simulate dispersal and quantify connectivity is lacking.

Here, we propose a simple three-step workflow that combines several
already-existing methods to assess connectivity using simulated dispersal
trajectories. In step one, we use integrated step selection functions to
parametrize a mechanistic movement model rendering dispersal behavior. In step
two, we apply the parametrized model as an individual-based movement model to
simulate dispersal trajectories. In step three, we combine simulated
trajectories into three complementary connectivity maps, each focusing on a
different aspect of landscape connectivity.

We showcase the application of the proposed workflow using data of dispersing
African wild dogs (\textit{Lycaon pictus}) and assess landscape connectivity
within the world's largest transboundary conservation area in Southern Africa.
We thereby shed light into dispersing wild dogs' habitat and movement
preferences, while also uncovering crucial dispersal corridors. With this
analysis, we demonstrate that simulations from integrated step selection
functions offer a simple, yet powerful alternative to traditional connectivity
modeling techniques.
\end{abstract}

\newpage
\begin{abstract}
%------------------------------------------------------------------------------
%  LONG VERSION
%------------------------------------------------------------------------------
\textbf{Long Version }
Dispersal of individuals is crucial for long-term species persistence and
depends on a sufficient degree of landscape connectivity. To date, researchers
primarily investigate connectivity using least-cost analysis and circuit theory,
both methods that make assumptions which are hardly met in reality. Least-cost
analysis assumes that animals move towards a known endpoint and are
knowledgeable about the most favorable route to reach it. Circuit theory
presumes a complete random walk and therefore fails to incorporate directional
persistence. These shortcomings can be overcome by spatio-temporally explicitly
simulating dispersal, yet a unified approach for such simulations is lacking.

Here, we present a simple three-step workflow to simulate dispersal and assess
connectivity starting from empirical GPS movement data. In step one, we use
integrated step selection functions to fit a mechanistic movement model
describing habitat and movement preferences of dispersers. In step two, we apply
the parameterized model to simulate individual dispersal trajectories. In step
three, we combine the simulated trajectories into three complementary
connectivity maps: a heatmap that highlights frequently traversed areas, a
betweenness map that pinpoints dispersal corridors, and a map of inter-patch
connectivity that indicates the presence and intensity of functional links
between habitat patches. As a case study, we analyse data on the endangered
African wild dog and assess landscape connectivity in the Kavango-Zambezi
Transfrontier Conservation Area, the world’s largest transboundary conservation
area.

Our results show that wild dogs preferably disperse with directional persistence
in the vicinity to water bodies and through areas with little human influence and
sparse woodland cover. Dispersal simulations and subsequent connectivity maps
revealed several dispersal hotspots and corridors across the extent of the
KAZA-TFCA. Connectivity between NPs inside the KAZA-TFCA was good, yet few
dispersers successfully moved from Zambia's NPs into neighboring areas.

We show that simulations from step-selection functions offer a simple yet
powerful alternative to traditional connectivity modeling techniques. In
contrast to traditional connect modeling techniques, simulations not only make
fewer unrealistic assumptions but also permit a more mechanistic understanding
of dispersal and landscape connectivity. This makes our workflow useful for a
variety of applications in ecological, evolutionary, and conservation research.

@@

% Importance of Dispersal & Connectivity
\subsection{Importance of Connectivity \& Connectivity Models}
Dispersal of individuals is a vital process that allows species to maintain
genetic diversity \citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012,
Baguette.2013}, rescue non-viable populations \citep{Brown.1977}, and colonize
unoccupied habitats \citep{Hanski.1999b, MacArthur.2001}. Importantly, the
ability to disperse depends on a sufficient degree of landscape connectivity
\citep{Fahrig.2003, Clobert.2012}, making the identification and protection of
dispersal corridors that promote connectivity a task of fundamental importance
\citep{Nathan.2008, Doerr.2011, Rudnick.2012}. The identification of dispersal
corridors not only necessitates a comprehensive understanding of the factors
that limit dispersal, but also an appropriate model to estimate connectivity
\citep{Baguette.2013, Vasudev.2015, Hofmann.2021}. To date, the two most
commonly used connectivity models are least-cost path analysis (LCPA;
\citealp{Adriaensen.2003}) and circuit theory (CT; \citealp{McRae.2006,
McRae.2008}), both graph-based methods that quantify conductance of the
landscape based on habitat permeability \citep{Zeller.2012}. However, both
approaches rest on assumptions that appear unsuitable for dispersers, which is
why simulating dispersal from individual-based movement models provides a
promising alternative approach to investigate connectivity \citep{Diniz.2019}.

% Issues with Both Methods
\subsection{Issues with Traditional Connectivity Models}
Traditional connectivity models make assumptions that are rarely met for
dispersers. LCPA, for instance, assumes that individuals move towards a
preconceived endpoint and choose a cost-minimizing route accordingly
\citep{Sawyer.2011, Abrahms.2017}. While this assumption may be fulfilled by
migrating animals, it is unlikely to hold for dispersers, as dispersers
typically move across unfamiliar territory, so the associated movement costs and
potential endpoints are unknown to them \citep{Koen.2014, Cozzi.2020}. CT, on
the contrary, posits that animals move according to a random walk, entailing
that autocorrelation between subsequent movements cannot be rendered
\citep{Diniz.2019}. For dispersers, however, autocorrelated movements are
regularly observed \citep{Cozzi.2020, Hofmann.2021}, implying that dispersal
trajectories are usually strongly directional. Moreover, graph-based methods are
unable to reflect the temporal dimension of dispersal, so that statements about
the expected duration for moving between habitats are impossible
\citep{Martensen.2017, Diniz.2019}.

% What about IBMMs?
\subsection{What about IBMMs?}
The shortcomings inherent to LCPA and CT can be overcome by simulating dispersal
from individual-based movement models (IBMMs) and converting simulated
trajectories into meaningful measures of connectivity \citep{Diniz.2019}). In
contrast to graph-based methods, IBMMs allow to explicitly simulate how
individuals move across and interact with the encountered landscape
\citep{Kanagaraj.2013, Clark.2015, Allen.2016, Hauenstein.2019, Zeller.2020}, as
well as to render potential interactions between movement behavior and habitat
conditions, thus shifting the focus from a structural to a more functional view
on connectivity \citep{Tischendorf.2000, Kanagaraj.2013, Hauenstein.2019}.
Furthermore, because simulations from IBMMs generate movement sequentially, the
temporal dimension of dispersal becomes explicit and allows to model
autocorrelation between subsequent movements \citep{Diniz.2019}. Finally,
simulations from IBMMs do not enforce connections towards preconceived
endpoints, thereby mitigating biases arising from misplaced endpoints. Despite
these advantages, a unified framework to simulate dispersal and assess
connectivity using IBMMs is lacking.

% Proposed Solution: Three-Step Workflow
\subsection{Proposed Solution: Three-Step Workflow}
Here, we draw on recent advancements in movement ecology and propose a simple
three-step workflow for simulating dispersal and assessing landscape
connectivity (\Cref{GraphicalAbstract}). In step one, we combine GPS movement
data of dispersing individuals with relevant habitat covariates to fit a
mechanistic movement model using integrated step selection functions (ISSFs,
\citealp{Avgar.2016}). ISSFs permit inference on the focal species' habitat
kernel (i.e. habitat preferences), its movement kernel (i.e. movement
preferences/capabilities), and potential interactions between the two
\citep{Avgar.2016, Fieberg.2021}. In step two, we simulate individual dispersal
trajectories using the parametrized movement from step one. Comparable
simulations have already been employed to estimate steady-state utilization
distributions of residents \citep{Potts.2013, Signer.2017} and to model
landscape connectivity, yet disregarding interdependencies between habitat and
movement kernels \citep{Clark.2015, Zeller.2020}. Finally, in step three, we
convert simulated trajectories into three complementary connectivity maps, each
highlighting a different aspect of connectivity. The set of maps includes (i) a
heatmap revealing areas that are frequently traversed by simulated dispersers
(e.g. \citealp{Hauenstein.2019, Zeller.2020}), (ii) a betweenness-map
delineating dispersal corridors and bottlenecks (e.g.
\citealp{BastilleRousseau.2018}), (iii) and a map of inter-patch connectivity,
depicting the presence and intensity of use of specific connections, as well as
the average dispersal duration required to connect corresponding habitat patches
(e.g. \citealp{Gustafson.1996, Kanagaraj.2013}).

% Introduction of the study species
\subsection{Case Study}
We showcase the application of the proposed workflow (\Cref{GraphicalAbstract})
using dispersal data collected on the endangered African wild dog
(\textit{Lycaon Pictus}). The African wild dog represents a highly mobile
species whose persistence hinges on a sufficient degree of landscape
connectivity. Once common throughout sub-Saharan Africa, this species has
disappeared from much of its historic range, largely due to human persecution,
habitat fragmentation, and disease outbreaks \citep{Woodroffe.2012}. Wild dogs
typically disperse in single-sex coalitions \citep{McNutt.1996, Behr.2020} and
are capable of dispersing several hundred kilometers \citep{DaviesMostert.2012,
Masenga.2016, Cozzi.2020}. Although previous studies have investigated
connectivity using LCPA \citep{Hofmann.2021} or CT \citep{Brennan.2020}, a more
comprehensive and mechanistic understanding of connectivity is missing for this
species (but see \citealp{Creel.2020}). With fewer than 6,000 free-ranging wild
dogs remaining in fragmented populations \citep{Woodroffe.2012}, reliable
information on landscape connectivity is essential for the conservation of this
endangered carnivore. Here, we use GPS data of 16 dispersing wild dogs
originating from a free-ranging population in northern Botswana to parametrize a
mechanistic movement model, which we then employed to simulate 80,000 dispersal
trajectories across the landscape of the world's largest transboundary
conservation area, the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA). We anticipated that simulations based on our three-step workflow
would overcome several of the above highlighted conceptual shortcomings of
traditional connectivity models and provide a more comprehensive view of
landscape connectivity

@@

\section{Methods}
\subsection{Study Area}
The study area centered at -17\degree 13'9''S, 23\degree 56'4''E
(\Cref{StudyArea}a) was located in southern Africa (\Cref{StudyArea}a) and
spanned more than 1.3 Mio. km\textsuperscript{2}, encompassing the entire
KAZA-TFCA (\Cref{StudyArea}b). The KAZA-TFCA comprises parts of Angola,
Botswana, Namibia, Zimbabwe, and Zambia and hosts a rich diversity of
landscapes, ranging from savannah to grassland and from dry to moist woodland
habitats. In its center lies the Okavango Delta, a dominant hydro-geographical
feature and the world's largest flood-pulsing inland delta. Large portions of
the KAZA-TFCA are part of designated national parks (NPs) and other protected
areas, yet considerable human influence through roads, agricultural sites, and
settlements remains.

@@

\subsection{Data Collection and Preparation}
\subsubsection{GPS Data}
We collected GPS movement data on 16 dispersing wild dogs (7 females and 9
males) between 2011 and 2019 from a free-ranging population in northern Botswana
(details on the data collection can be found in \cite{Cozzi.2020} and
\cite{Hofmann.2021}). Because behavior during dispersal is more pertinent to
landscape connectivity than behavior during residence \citep{Elliot.2014,
Abrahms.2017}, we discarded data collected when individuals were resident. We
determined the exact timing of emigration and settlement using direct field
observations and through visual inspection of the net squared displacement (NSD)
metric. NSD measures the squared Euclidean distance of a GPS relocation to a
reference point \citep{Borger.2012}, which we set to the center of each
dispersers' natal home range. Thus, dispersal was deemed to have started once an
individual left its natal home range and ended once the NSD metric remained
constant, indicating settlement. During dispersal, GPS collars recorded a fix
every 4 hours and regularly transmitted data over the Iridium satellite system.
To ensure regular timespans between GPS fixes, we removed any fix that was not
successfully obtained on the aspired 4-hourly schedule (allowing for a tolerance
of \( \pm \) 15 minutes). We then converted the remaining fixes (n = 4'169) into
steps, where each step represented the straight-line movement between two
consecutive GPS fixes \citep{Turchin.1998}.

\subsubsection{Habitat Covariates}
We represented the physical landscape in our study area by the habitat
covariates \textit{water-cover, distance-to-water, woodland-cover,
shrub/grassland-cover, and human-influence}. To correctly render seasonality in
water-cover, we applied a remote sensing algorithm that enabled us to obtain
weekly updated raster-layers for water-cover and corresponding layers for
distance-to-water from MODIS satellite imagery \citealp{Wolski.2017,
Hofmann.2021}. This algorithm is now implemented in the \textsf{floodmapr}
package which is available on GitHub
(\url{https://github.com/DavidDHofmann/floodmapr}). Using the frequently updated
floodmpas we were able to correctly assign the most appropriate covariate layer
to each observed movement step. To ensure a consistent resolution across habitat
covariates, we coarsened or interpolated all layers to a resolution of 250 m x
250 m. A detailed description of how we prepared each habitat covariate is
provided in \cite{Hofmann.2021}.

Besides habitat covariates, we computed movement metrics that we used as
movement covariates in ISSF models \citep{Avgar.2016, Fieberg.2021}. Movement
metrics were calculated for each step and included the step length
(\textsf{sl}), its natural logarithm (\textsf{log(sl)}), and the cosine of the
relative turning angle (\textsf{cos(ta)}). Moreover, we created the binary
variable \textsf{LowActivity} indicating whether a step was realized during
periods of low wild dog activity (09:00 to 17:00 local time) or high wild dog
activity (17:00 to 09:00 local time, \citealp{Cozzi.2012}). We performed all
data preparations, spatial computations, and statistical analysis in R, version
3.6.6 \citep{R.2020}. Some helper functions were written in {\tt C++} and
imported into {\tt R} using the {\tt Rcpp} package \citep{Eddelbuettel.2011,
Eddelbuettel.2013}.

\subsection{Step 1 - Movement Model}
We used ISSFs to parametrize a mechanistic movement model for dispersing wild
dogs \citep{Avgar.2016}. More specifically, we paired each realized step with 24
random steps, so that a realized step plus its 24 random steps formed a
25-step-stratum that received a unique identifier. As suggested by
\cite{Avgar.2016}, we generated random steps by sampling random turning angles
from a uniform distribution (\(-\pi, +\pi\)) (which is equivalent to a von Mises
distribution with location and concentration parameters; \(\mu = \kappa = 0\))
and step lengths from a gamma distribution that was fitted to realized steps
(scale \(\theta\) = 6'308 and shape \(k\) = 0.37). Along each step, we extracted
and averaged values from the habitat covariate layers using the {\tt velox}
package \citep{Hunziker.2021} and calculated the movement metrics \textsf{sl},
\textsf{log(sl)}, and \textsf{cos(ta)} for each realized and random step. To
facilitate model convergence, we standardized all continuous covariates to a
mean of zero and a standard deviation of one. Since correlation among covariates
was low (\(|r| < 0.6\); \citealp{Latham.2011}), we retained all of them for
modeling.

To contrast realized steps (scored 1) and random steps (scored 0), we assumed
that animals assigned a selection score \(w(x)\) of the exponential form to each
step (\ref{EQ2}; \citealp{Fortin.2005}). \(w(x)\) depended on the step's
associated covariates (\(x_1, x_2, ..., x_n\)) and on the animal's preferences
(i.e. relative selection strengths; \citealp{Avgar.2017}) towards these
covariates (\(\beta_1, \beta_2, ..., \beta_n\)):

@@

\noindent The probability of a step being realized was then contingent on the
step's selection score, as well as on the selection scores of all other step in
the same stratum:

@@

\noindent To estimate preferences (i.e. the \(\beta\)-coefficients), we used
mixed effects conditional logistic regression analysis \citep{Muff.2020} that we
implemented using the r-package {\tt glmmTMB} \citep{Brooks.2017}. To capitalize
on the poisson trick proposed by \cite{Muff.2020}, we defined random intercepts
for each stratum and fixed the intercept variance to an arbitrary high value of
\(10 ^ 6\). We also used disperser identity to model random slopes for all
covariates.

>>>>

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{A Three Steps Workflow for Simulating Dispersal Trajectories and
Assessing Landscape Connectivity: A Case Study on the Endangered African Wild
Dog}

@@

% Current Date
% \date{\today}

@@

\textsuperscript{2} Botswana Predator Conservation Program, Private Bag 13,
Maun, Botswana.

@@

\textbf{Running Title:} Simulating Wild Dog Dispersal Trajectories to Assess
Landscape Connectivity

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
% \newpage
% \begin{abstract}
% %------------------------------------------------------------------------------
% % Short Version
% %------------------------------------------------------------------------------
% \textbf{Short Version }
% The ability to disperse is contingent on a sufficient degree of landscape
% connectivity, which is why the identification and preservation of movement
% corridors that promote connectivity has become a task of extraordinary
% importance. Currently, ecologists rely on least-cost analysis or circuit theory
% to investigate connectivity, although both methods make several assumptions that
% are hardly met in reality. To address these issues, simulations from
% individual-based movement models have been proposed, yet a unified framework to
% simulate dispersal and quantify connectivity is lacking.
%
% Here, we propose a simple three-step workflow that combines several
% already-existing methods to assess connectivity using simulated dispersal
% trajectories. In step one, we use integrated step selection functions to
% parametrize a mechanistic movement model rendering dispersal behavior. In step
% two, we apply the parametrized model as an individual-based movement model to
% simulate dispersal trajectories. In step three, we combine simulated
% trajectories into three complementary connectivity maps, each focusing on a
% different aspect of landscape connectivity.
%
% We showcase the application of the proposed workflow using empirical data of
% dispersing African wild dogs (\textit{Lycaon pictus}) and assess landscape
% connectivity within the world's largest transboundary conservation area in
% Southern Africa. We thereby shed light into dispersing wild dogs' habitat and
% movement preferences, while also uncovering crucial dispersal corridors. With
% this analysis, we demonstrate that simulations from integrated step selection
% functions offer a simple, yet powerful alternative to traditional connectivity
% modeling techniques.
% \end{abstract}

\newpage
\begin{abstract}
%------------------------------------------------------------------------------
%  LONG VERSION
%------------------------------------------------------------------------------
Dispersal of individuals contributes to long-term population persistence, yet
requires a sufficient degree of landscape connectivity. To date, connectivity
has mainly been investigated using least-cost analysis and circuit theory, two
methods that make assumptions that are hardly applicable during dispersal.
Least-cost analysis assumes that animals move towards a known endpoint and are
knowledgeable about the most favorable route to reach it. Circuit theory
presumes a complete random walk and fails to incorporate directional
persistence. While these assumptions can be relaxed by simulating dispersal
movements spatio-temporally explicitly across the landscape of interest, a
unified approach for such simulations is lacking.

Here, we present a simple three-step workflow to simulate dispersal movements
and to assess connectivity starting from empirical GPS movement data. In step
one, we use integrated step selection functions to fit a mechanistic movement
model describing habitat and movement preferences of dispersing individuals. In
step two, we apply the parameterized movement model to simulate dispersal
trajectories. In step three, we derive three complementary connectivity maps: a
heatmap that highlights frequently traversed areas, a betweenness map that
pinpoints dispersal corridors, and a map of inter-patch connectivity that
indicates the presence and intensity of functional links between habitat
patches. As a case study, we applied the workflow to GPS data collected during
dispersal of the endangered African wild dog (\textit{Lycaon pictus}) inhabiting
the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA).

Results from the case study show that wild dogs preferably disperse with
directional persistence in the vicinity to water bodies, through areas with
little human influence and sparse woodland cover. Dispersal simulations and
subsequent connectivity maps revealed several dispersal hotspots and corridors
across the extent of the KAZA-TFCA. Connectivity between NPs inside the
KAZA-TFCA was good, yet few dispersers successfully moved from Zambia's NPs into
neighboring areas. By rendering interactions between movement preferences and
habitat preferences, our movement model substantially outperformeed a previous
model that omitted such interactions. While we simulated 80,000 dispersal
trajectories in total, we found that metrics of connectivity only marginally
changed beyond 10,500 simulations, suggesting that fewer simulations would have
sufficed.

Ultimately, we show that a simulation-based approach that leverages on
step-selection functions offers a simple yet powerful alternative to traditional
connectivity modeling techniques. Such an approach not only makes fewer
biologically unrealistic assumptions but also permits a more mechanistic
understanding of dispersal and landscape connectivity. Our workflow is thus
useful for a variety of applications in ecological, evolutionary, and
conservation research.

@@

% Importance of Dispersal & Connectivity
\subsection{Importance of Connectivity \& Connectivity Models}
Dispersal of individuals is a vital process that allows species to maintain
genetic diversity \citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012,
Baguette.2013}, rescue non-viable populations \citep{Brown.1977}, and colonize
unoccupied habitats \citep{Hanski.1999b, MacArthur.2001}. Importantly, the
ability to disperse depends on a sufficient degree of landscape connectivity
\citep{Fahrig.2003, Clobert.2012}, making the identification and protection of
dispersal corridors that promote connectivity a task of fundamental importance
\citep{Nathan.2008, Doerr.2011, Rudnick.2012}. Identifying dispersal corridors
not only necessitates a comprehensive understanding of the factors that limit
dispersal movements, but also an appropriate model to estimate connectivity
\citep{Baguette.2013, Vasudev.2015, Hofmann.2021}. To date, the most commonly
used models to assess connectivity are least-cost path analysis (LCPA;
\citealp{Adriaensen.2003}) and circuit theory (CT; \citealp{McRae.2006,
McRae.2008}). Unfortunately, both models rest on assumptions that appear
unsuitable for dispersers, calling for the development of alternative
approaches. One promising alternative to investigate dispersal and assess
connectivity is to simulate dispersal from individual-based movement models
\citep{Diniz.2019}.

% Issues with Both Methods
\subsection{Issues with Traditional Connectivity Models}
Traditional connectivity models make assumptions that are rarely met for
dispersers. LCPA, for instance, assumes that individuals move towards a
preconceived endpoint and choose a cost-minimizing route accordingly
\citep{Sawyer.2011, Abrahms.2017}. While this assumption may be justifiable for
migrating animals, it is unlikely to hold for dispersers, as dispersers
typically move across unfamiliar territory towards an unknown endpoint
\citep{Koen.2014, Cozzi.2020}. CT, on the contrary, posits that animals move
according to a random walk, entailing that autocorrelation between subsequent
movements cannot be rendered \citep{Diniz.2019}. For dispersers, however,
autocorrelated movements are regularly observed \citep{Cozzi.2020,
Hofmann.2021}, implying that dispersal trajectories are usually strongly
directional. Moreover, because both models require static permeability or
resistance surfaces as input, they are unable to reflect the temporal dimension
of dispersal, meaning that statements about the expected duration for moving
between habitat patches are impossible \citep{Martensen.2017, Diniz.2019}.

% What about IBMMs?
\subsection{What about IBMMs?}
The shortcomings inherent to LCPA and CT can be overcome by simulating dispersal
from individual-based movement models (IBMMs) and by converting simulated
trajectories into meaningful measures of connectivity \citep{Diniz.2019}). In
contrast to LCPA and CT, IBMMs allow to explicitly simulate how individuals move
across and interact with the encountered landscape \citep{Kanagaraj.2013,
Clark.2015, Allen.2016, Hauenstein.2019, Zeller.2020}, as well as to render
potential interactions between movement behavior and habitat conditions
\citep{Avgar.2016}. This strictly shifts the focus from a structural to a more
functional view on connectivity \citep{Tischendorf.2000, Kanagaraj.2013,
Hauenstein.2019}. Furthermore, IBMMs generate movement sequentially, i.e. they
generate a series of steps, the temporal dimension of dispersal movements
becomes explicit and allows to model autocorrelation between consecutive steps
\citep{Diniz.2019}. Finally, simulations from IBMMs do not enforce movement or
connections towards preconceived endpoints, thereby preventing biases arising
from misplaced endpoints. Despite these advantages, a unified framework to
simulate dispersal and assess connectivity using IBMMs is lacking.

% Proposed Solution: Three-Step Workflow
\subsection{Proposed Solution: Three-Step Workflow}
Here, we propose a simple three-step workflow for simulating dispersal and
assessing landscape connectivity (\Cref{GraphicalAbstract}). In step one, we
combine GPS movement data of dispersing individuals with relevant habitat
covariates to fit a mechanistic movement model using integrated step selection
functions (ISSFs, \citealp{Avgar.2016}). ISSFs allow inference on the study
species' habitat kernel (i.e. habitat preferences), its movement kernel (i.e.
movement preferences/capabilities), and potential interactions between the two
\citep{Avgar.2016, Fieberg.2021}. In step two, we simulate individual dispersal
trajectories using the parametrized movement model. Comparable simulations have
already been applied to estimate steady-state utilization distributions of
resident individuals \citep{Potts.2013, Signer.2017} and to model landscape
connectivity, yet disregarding interdependencies between habitat and movement
kernels \citep{Clark.2015, Zeller.2020}. Finally, in step three, we convert
simulated trajectories into three complementary connectivity maps, each
highlighting a different aspect of connectivity. The set of maps includes (i) a
heatmap revealing areas that are frequently traversed by dispersers (e.g.
\citealp{Hauenstein.2019, Zeller.2020}), (ii) a betweenness-map delineating
dispersal corridors and bottlenecks (e.g. \citealp{BastilleRousseau.2018}),
(iii) and a map of inter-patch connectivity, depicting the presence and
intensity of specific connections, as well as the average dispersal duration
required to connect corresponding habitat patches (e.g. \citealp{Gustafson.1996,
Kanagaraj.2013}).

% Introduction of the study species
\subsection{Case Study}
We showcase the application of the proposed workflow (\Cref{GraphicalAbstract})
using GPS movement data collected on dispersing individuals of the endangered
African wild dog (\textit{Lycaon Pictus}). The African wild dog is a highly
mobile species whose population persistence heavily relies on the availability
of large, natural or semi-natural landscapes and a sufficient degree of
connectivity among remaining subpopulations. Once common throughout sub-Saharan
Africa, this species has disappeared from much of its historic range, largely
due to human persecution, habitat fragmentation, and disease outbreaks
\citep{Woodroffe.2012}. Wild dogs typically disperse in single-sex coalitions
\citep{McNutt.1996, Behr.2020} and are capable of dispersing several hundred
kilometers \citep{DaviesMostert.2012, Masenga.2016, Cozzi.2020}. Although
previous studies have investigated connectivity using LCPA \citep{Hofmann.2021}
or CT \citep{Brennan.2020}, a more comprehensive and mechanistic understanding
of connectivity is missing for this species (but see \citealp{Creel.2020}). With
fewer than 6,000 free-ranging wild dogs remaining in fragmented subpopulations
\citep{Woodroffe.2012}, reliable information on landscape connectivity is
essential for the conservation of this endangered carnivore. Here, we use GPS
data of 16 dispersing wild dogs originating from a free-ranging population in
northern Botswana to parametrize a mechanistic movement model, which we then
employed to simulate 80,000 dispersal trajectories across the landscape of the
world's largest transboundary conservation area, the Kavango-Zambezi
Transfrontier Conservation Area (KAZA-TFCA). We anticipated that simulations
based on our three-step workflow would overcome several of the above highlighted
conceptual shortcomings of traditional connectivity models and provide a more
comprehensive view on dispersal behavior and landscape connectivity.

@@

\section{Methods}
\subsection{Study Area}
Our simulation of dispersal trajectories and assessment of connectivity spanned
across the entire Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA),
an area of approximately 1.3 Mio. km\textsuperscript{2} (\Cref{StudyArea}a and
b). The KAZA-TFCA is the world's largest transboundary conservation area and
comprises parts of Angola, Botswana, Namibia, Zimbabwe, and Zambia, thus hosting
a rich diversity of landscapes, ranging from savannah to grassland and from dry
to moist woodland habitats. In its center lies the Okavango Delta, a dominant
hydro-geographical feature and the world's largest flood-pulsing inland delta.
Large portions of the KAZA-TFCA are formally protected in the form of national
parks (NPs) or other protected areas, yet considerable human-dominated
landscapes (e.g. roads, agricultural sites, and settlements) remain.

@@

\subsection{Data Collection and Preparation}
\subsubsection{GPS Data}
We collected GPS movement data on 16 coalitions of dispersing African wild dogs
(7 female and 9 male coalitions) between 2011 and 2019 from a free-ranging
population in northern Botswana (details on the data collection can be found in
\cite{Cozzi.2020} and \cite{Hofmann.2021}). To delineate periods of dispersal,
we determined the exact timing of emigration from the natal pack and settlement
in a new territory using direct field observations and through visual inspection
of the net squared displacement (NSD) metric. The NSD metric measures the
squared Euclidean distance of a GPS relocation to a reference point
\citep{Borger.2012}, which we set to the center of the territory of each
dispersers' natal pack. Thus, dispersal was deemed to have started once an
individual left its natal territory and ended once the NSD metric remained
constant, indicating settlement. Because behavior during dispersal is more
pertinent to landscape connectivity than behavior during residence
\citep{Elliot.2014, Abrahms.2017}, we only considered data collected during
dispersal for our analysis. During dispersal, GPS collars recorded a fix every 4
hours and regularly transmitted data over the Iridium satellite system. To
ensure regular time intervals between GPS fixes, we removed any fixes that were
not successfully obtained at the desired 4-hour schedule (allowing for a
tolerance of \( \pm \) 15 minutes). We then converted the fixes (n = 4'169) into
steps, where each step represented the straight-line movement between two
consecutive GPS fixes \citep{Turchin.1998}.

\subsubsection{Habitat Covariates}
We represented the physical landscape in our study area by the habitat
covariates \textit{water-cover, distance-to-water, woodland-cover,
shrub/grassland-cover, and human-influence}. To render the seasonal dynamics in
water-cover of major water bodies for the extent of the Okavango Delta, we
applied an algorithm that enabled us to obtain weekly updated raster-layers for
water-cover and distance-to-water from MODIS satellite imagery
\citep{Wolski.2017, Hofmann.2021}. This algorithm is now implemented in the
\textsf{floodmapr} package (available on GitHub;
\url{https://github.com/DavidDHofmann/floodmapr}). To ensure a consistent
resolution across habitat covariates, we coarsened or interpolated all layers to
a resolution of 250 m x 250 m. A detailed description of how we prepared each
habitat covariate is provided in \cite{Hofmann.2021}.

Besides habitat covariates, we computed movement metrics that we used as
movement covariates in the ISSF models \citep{Avgar.2016, Fieberg.2021}.
Specifically, we computed for each step the step length (\textsf{sl}), its
natural logarithm (\textsf{log(sl)}), and the cosine of the relative turning
angle (\textsf{cos(ta)}). Moreover, we created the binary variable
\textsf{LowActivity}, indicating whether a step was realized during periods of
low wild dog activity (09:00 to 17:00 local time) or high wild dog activity
(17:00 to 09:00 local time, following \citealp{Cozzi.2012}). We performed all
data preparations, spatial computations, and statistical analysis in R, version
3.6.6 \citep{R.2020}. Some helper functions were written in {\tt C++} and
imported into {\tt R} using the {\tt Rcpp} package \citep{Eddelbuettel.2011,
Eddelbuettel.2013}.

\subsection{Step 1 - Movement Model}
We used ISSFs \citep{Avgar.2016} to parametrize a mechanistic movement model for
dispersing wild dogs. More specifically, we paired each realized (i.e. observed)
step with 24 random steps, so that a realized step plus its 24 random steps
formed a 25-step-stratum that received a unique identifier. As suggested by
\cite{Avgar.2016}, we generated random steps by sampling random turning angles
from a uniform distribution (\(-\pi, +\pi\)) (which is equivalent to a von Mises
distribution with location and concentration parameters; \(\mu = \kappa = 0\))
and step lengths from a gamma distribution that was fitted to realized steps
(scale \(\theta\) = 6'308 and shape \(k\) = 0.37).

Along each realized and random step, we extracted values from the habitat
covariate layers using the {\tt velox} package \citep{Hunziker.2021} and we
computed averages of each covariate along the steps. We further calculated the
movement metrics \textsf{sl}, \textsf{log(sl)}, and \textsf{cos(ta)} for each
step. To facilitate model convergence, we standardized all continuous covariates
to a mean of zero and a standard deviation of one. Correlations among covariates
were low (\(|r| < 0.6\); \citealp{Latham.2011}), so we retained all of them for
modeling.

To contrast realized steps (scored 1) and random steps (scored 0), we assumed
that animals assigned a selection score \(w(x)\) to each step (\ref{EQ1};
\citealp{Fortin.2005}). \(w(x)\) depended on the step's associated covariates
(\(x_1, x_2, ..., x_n\)) and on the animal's preferences (i.e. relative
selection strengths; \citealp{Avgar.2017}) towards these covariates (\(\beta_1,
\beta_2, ..., \beta_n\)):

@@

\noindent The probability of a step \(i\) being realized was then contingent on
the step's selection score, as well as on the selection scores of all other step
in the same stratum:

@@

\noindent To estimate preferences (i.e. the \(\beta\)-coefficients), we used
mixed effects conditional logistic regression analysis \citep{Muff.2020} that we
implemented using the r-package {\tt glmmTMB} \citep{Brooks.2017}. The method
introduced by \cite{Muff.2020} allows to model random slopes, yet requires to
fix the variance of the stratum specific intercept to a large value. Hence we
fixed the stratum specific intercept variance to an arbitrary high value of \(10
^ 6\). We used disperser identity to model random slopes for all covariates.

The covariate structure of the movement model was based on a habitat selection
model that was previously developed for dispersing wild dogs (hereafter referred
to as \textit{based model}, \citealp{Hofmann.2021}). In the base model, no
interactions among habitat covariates and movement covariates were considered.
Hence, we expanded the base model and allowed for interactions between all
movement covariates and habitat covariates, thus reflecting that movement
behavior may depend on habitat conditions (details in Appendix A1). To determine
the most parsimonious movement model among model candidates, we ran stepwise
forward model selection based on Akaike's Information Criterion (AIC,
\citealp{Burnham.2002}). Finally, we validated the predictive power of the most
parsimonious model using k-fold cross-validation for case-control studies as
described in \cite{Fortin.2009}. This validation proves a significant prediction
in case the Spearman rank correlation of predicted step-ranks and associated
frequencies under the movement model is significantly greater than under the
assumption of random preferences (details in Appendix A2).
